---
title: Visualisasi Maps (Leaflet)
author: Team Laboratorium Biro TI
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: united
    highlight: espresso
    fig_width: 10
    fig_height: 5
    df_print: paged
    code_folding: show
    #mathjax: null
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

options(scipen = 9999)
```

# Init Data

## Library

```{r}
# General
library(tidyverse)
library(glue)
library(scales)

# Read Json
library(jsonlite)

# Maps
library(leaflet)
library(sf)
library(htmltools)

# Read xlsx
library(readxl)
```

## Load Data
```{r}
mapPoly <-  read_sf("./data/prov.bps.json")
mapCases <- fromJSON("./data/raw_covid.json")
mapPenduduk <- read_excel("./data/raw_maps.xlsx", sheet = 1)
mapikf <- read_excel("./data/raw_maps.xlsx", sheet = 3)
```

# Maps

## Basemap

```{r}
basemap <- leaflet() %>%
  addTiles() %>% 
  
  # Options Maps
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  # Set View Map (Center)
  fitBounds(95.240772,-10.919438,141.019400,5.649667) %>% 
  
  # Add Custom Button (Home)
  addEasyButton(easyButton(
     icon="fa-globe", title="Home", position = "topleft",
     onClick=JS("function(btn, map){ map.setZoom(4); map.setView([-2.548926,118.0148634],4); }")))

basemap
```

## Marker

### Simple Marker

```{r}
basemap %>% 
  addCircleMarkers(
    # data
    data = mapCases, 
    
    #Latitude
    lat = ~ lokasi$lat, 
    
    #Longitude
    lng = ~ lokasi$lon
  )
```

### Custom Marker
```{r}
basemap %>% 
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    
    # Weigth Line
    weight = 1,
    
    # Weight Circle
    radius = ~(jumlah_kasus)^(1/4.5),
    
    # Color
    color = "#cc4c02"
  )
```


## Labels

### Simple Labels

```{r}
basemap %>% 
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_kasus)^(1/4.5),
    color = "#cc4c02",
    
    # Add Label
    label = paste(mapCases$key,mapCases$jumlah_kasus,sep = ": ")
  )
```

### Custom Labels

```{r}
basemap %>% 
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_kasus)^(1/4.5),
    color = "#cc4c02",
    
    # Format Numeric
    label = paste(mapCases$key,comma(mapCases$jumlah_kasus,accuracy = 1), sep = ": ")
  )
```

** Make Function **
```{r}
labels_map <-
  glue("<b>{mapCases$key}</b><br>
       Total Cases : {comma(mapCases$jumlah_kasus,accuracy = 1)}<br>
       Recover Cases : {comma(mapCases$jumlah_sembuh,accuracy = 1)}<br>
       Death Cases : {comma(mapCases$jumlah_meninggal,accuracy = 1)}") %>%
  lapply(htmltools::HTML)
```


```{r}
basemap %>% 
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_kasus)^(1/4.5),
    color = "#cc4c02",
    
    # Make Function
    label = labels_map,
    
    # Style Label
    labelOptions = labelOptions(
                     style = list("font-weight" = "normal", padding = "3px 8px", "color" = "#cc4c02"),
                     textsize = "14px", direction = "auto")
  )
```

## Layer Markers

### Add Layer Markers
```{r}
basemap %>% 
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_kasus)^(1/4.5),
    color = "#cc4c02",
    label = labels_map,
    labelOptions = labelOptions(
                     style = list("font-weight" = "normal", padding = "3px 8px", "color" = "#cc4c02"),
                     textsize = "14px", direction = "auto")
  ) %>% 
  
  # Add Marker
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_meninggal)^(1/4.5),
    label = labels_map,
    labelOptions = labelOptions(
                     style = list("font-weight" = "normal", padding = "3px 8px", "color" = "#cc4c02"),
                     textsize = "14px", direction = "auto")
  )
```


### Add Group Marker

```{r}
basemap %>% 
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_kasus)^(1/4.5),
    color = "#cc4c02",
    label = labels_map,
    labelOptions = labelOptions(
                     style = list("font-weight" = "normal", padding = "3px 8px", "color" = "#cc4c02"),
                     textsize = "14px", direction = "auto"),
    # Add Group
    group = "Kasus"
  ) %>% 
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_meninggal)^(1/4.5),
    label = labels_map,
    labelOptions = labelOptions(
                     style = list("font-weight" = "normal", padding = "3px 8px", "color" = "#cc4c02"),
                     textsize = "14px", direction = "auto"),
    # Add Group
    group = "Meninggal"
  ) %>% 
  
  # Add Group
  addLayersControl(
    position = "bottomright",
    baseGroups = c("Kasus", "Meninggal"),
    options = layersControlOptions(collapsed = FALSE)) %>% 
  hideGroup(c("Meninggal"))
```

## Polygon

### Init Data

```{r}
map_merge <- merge(
  mapPoly, 
  select(mapPenduduk, c(nama_provinsi, jumlah_penduduk)), 
  by.x = "provinsi", 
  by.y = "nama_provinsi")

map_merge
```

### Single Polygon

#### Simple Legend
```{r}
basemap %>%
  # Polygon
  addPolygons(data = map_merge,
              group = 'Penduduk',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~colorBin("YlOrRd", domain = map_merge$jumlah_penduduk, bins = 6)(jumlah_penduduk),
              label = ~provinsi) %>% 
  
  # Legend
  addLegend(
    "bottomleft",
    pal = colorBin("YlOrRd", domain = map_merge$jumlah_penduduk, bins = 6),
    values = map_merge$jumlah_penduduk, title='Penduduk', group = 'Penduduk')
```

#### Custom Legend Scale

```{r}
bins <- c(500000, 1000000, 3000000, 5000000, 10000000, 20000000, 30000000,Inf)
pal <- colorBin("YlOrRd", domain = map_merge$jumlah_penduduk, bins = bins)
```

```{r}
basemap %>%
  addPolygons(data = map_merge,
              group = 'Penduduk',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal(jumlah_penduduk),
              label = ~provinsi) %>% 
  addLegend(
    "bottomleft",
    pal = pal,
    values = map_merge$jumlah_penduduk, title='Penduduk', group = 'Penduduk')
```


### Multiple Polygon

#### Merge Poligon

```{r}
map_merge_poly <- 
  merge(map_merge, mapikf %>% select(-kode_provinsi), by.x = "provinsi", by.y = "nama_provinsi")

map_merge_poly
```

```{r}
bins_ikf <- c(.0100, .1000, .200, .3000, .5000,Inf)
pal_ikf <- colorBin("YlGnBu", domain = map_merge_poly$t_2020, bins = bins_ikf)
```

```{r}
basemap %>%
  addPolygons(data = map_merge_poly,
              group = 'Penduduk',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal(jumlah_penduduk),
              label = ~provinsi) %>% 
  addLegend(
    "bottomleft",
    pal = pal,
    values = map_merge$jumlah_penduduk, title='Penduduk', group = 'Penduduk') %>% 
  
  # New Polygon
  addPolygons(data = map_merge_poly,
              group = 'IKF',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal_ikf(t_2020),
              label = ~provinsi) %>% 
  addLegend(
    "bottomleft",
    pal = pal_ikf,
    values = map_merge_poly$t_2020, title='IKF', group = 'IKF')
```


#### Add Group

```{r}
basemap %>%
  addPolygons(data = map_merge_poly,
              group = 'Penduduk',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal(jumlah_penduduk),
              label = ~provinsi) %>% 
  addLegend(
    "bottomleft",
    pal = pal,
    values = map_merge$jumlah_penduduk, title='Penduduk', group = 'Penduduk') %>% 
  
  # New Polygon
  addPolygons(data = map_merge_poly,
              group = 'IKF',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal_ikf(t_2020),
              label = ~provinsi) %>% 
  addLegend(
    "bottomleft",
    pal = pal_ikf,
    values = map_merge_poly$t_2020, title='IKF', group = 'IKF') %>% 
  
  # Add Group
  addLayersControl(
    position = "bottomright",
    overlayGroups  = c("Penduduk", "IKF"),
    options = layersControlOptions(collapsed = FALSE)) %>% 
  hideGroup(c("IKF"))
```


# Latihan

## Instruksi

- Gabungkan Marker dan Polygon Menjadi Satu Maps

_ Tambahkan Judul Peta


## Pembahasan

### Merge Data

```{r}
map_merge_poly <- 
  merge(mapPoly, select(mapPenduduk, c(nama_provinsi, jumlah_penduduk)), by.x = "provinsi", by.y = "nama_provinsi") %>% 
  merge(mapikf %>% select(-kode_provinsi), by.x = "provinsi", by.y = "nama_provinsi")

map_merge_poly
```

### Map Ikf

```{r}
merge_basemap <- basemap %>% 
  addPolygons(data = map_merge_poly,
              group = 'IKF',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal_ikf(t_2020),
              label = ~provinsi) %>% 
  addLegend(
    "bottomleft",
    pal = pal_ikf,
    values = map_merge_poly$t_2020, title='IKF', group = 'IKF')

merge_basemap
```

\
**Custom Label**

```{r}
# Add Simbol
up_arrow <- "<span style=\"color:green\">&#9650;</span>"
down_arrow <- "<span style=\"color:red\">&#9660;</span>"

# Custom Label
labels_map_ikf_prov <-
  glue("<span style=font-size:14px;><b>{str_to_title(mapikf$nama_provinsi)}</b><br>
       2020 : {comma(mapikf$t_2020)} {ifelse(mapikf$t_2020 > mapikf$t_2019, up_arrow, down_arrow)}<br>
       2019 : {comma(mapikf$t_2019)} {ifelse(mapikf$t_2019 > mapikf$t_2018, up_arrow, down_arrow)}<br>
       2018 : {comma(mapikf$t_2018)} {ifelse(mapikf$t_2018 > mapikf$t_2017, up_arrow, down_arrow)}<br>
       2017 : {comma(mapikf$t_2017)} {ifelse(mapikf$t_2017 > mapikf$t_2016, up_arrow, down_arrow)}<br>
       2016 : {comma(mapikf$t_2016)} {ifelse(mapikf$t_2016 > mapikf$t_2015, up_arrow, down_arrow)}<br>
       2015 : {comma(mapikf$t_2015)} {ifelse(mapikf$t_2015 > mapikf$t_2014, up_arrow, down_arrow)}<br>
       2014 : {comma(mapikf$t_2014)} {ifelse(mapikf$t_2014 > mapikf$t_2013, up_arrow, down_arrow)}<br>
       2013 : {comma(mapikf$t_2013)}</span>") %>%
  lapply(htmltools::HTML)
```

\
**Map Ikf with Custom Label**

```{r}
merge_basemap <- basemap %>% 
  addPolygons(data = map_merge_poly,
              group = 'IKF',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal_ikf(t_2020),
              label = ~labels_map_ikf_prov) %>% 
  addLegend(
    "bottomleft",
    pal = pal_ikf,
    values = map_merge_poly$t_2020, title='IKF', group = 'IKF')

merge_basemap
```



### Merge All 

```{r}
merge_basemap <- basemap %>% 
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_kasus)^(1/4.5),
    color = "#cc4c02",
    label = labels_map,
    labelOptions = labelOptions(
                     style = list("font-weight" = "normal", padding = "3px 8px", "color" = "#cc4c02"),
                     textsize = "14px", direction = "auto"),
    group = "Kasus"
  ) %>% 
  
  addCircleMarkers(
    data = mapCases, 
    lat = ~ lokasi$lat, 
    lng = ~ lokasi$lon,
    weight = 1,
    radius = ~(jumlah_meninggal)^(1/4.5),
    label = labels_map,
    labelOptions = labelOptions(
                     style = list("font-weight" = "normal", padding = "3px 8px", "color" = "#cc4c02"),
                     textsize = "14px", direction = "auto"),
    group = "Meninggal"
  ) %>%
  
  addPolygons(data = map_merge_poly,
              group = 'Penduduk',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal(jumlah_penduduk),
              label = ~provinsi) %>% 
  addLegend(
    "bottomleft",
    pal = pal,
    values = map_merge$jumlah_penduduk, title='Penduduk', group = 'Penduduk') %>% 
  

  addPolygons(data = map_merge_poly,
              group = 'IKF',
              weight=1, opacity = 1.0,color = 'green',
              fillOpacity = 0.6, smoothFactor = 0.5,
              fillColor = ~pal_ikf(t_2020),
              label = ~labels_map_ikf_prov) %>% 
  addLegend(
    "bottomleft",
    pal = pal_ikf,
    values = map_merge_poly$t_2020, title='IKF', group = 'IKF') %>% 
  
  
  addLayersControl(
    position = "bottomright",
    baseGroups = c("Kasus", "Meninggal"),
    overlayGroups  = c("Penduduk", "IKF"),
    options = layersControlOptions(collapsed = FALSE)) %>% 
  hideGroup(c("Meninggal","IKF"))

merge_basemap
```


### Change Group

```{r}
merge_basemap <- merge_basemap %>% 
  addLayersControl(
    position = "bottomright",
    overlayGroups  = c("Kasus", "Meninggal", "Penduduk", "IKF"),
    options = layersControlOptions(collapsed = FALSE)) %>% 
  hideGroup(c("Meninggal","Penduduk", "IKF"))

merge_basemap
```


### Add title

```{r}
tag.map.title <- tags$style(HTML("
  .leaflet-control.map-title { 
    transform: translate(100%,20%);
    position: absolute !important;
    width: 300px;
    text-align: center;
    padding-left: 2px; 
    padding-right: 2px; 
    background: #f5e23d;
    font-weight: bold;
    font-size: 16px;
  }
"))

title <- tags$div(
  tag.map.title, HTML("Sebaran Covid, Penduduk dan IKF")
) 
```


## Final Map

```{r}
merge_basemap %>% 
  addControl(title, className="map-title")
```


# Referensi

- https://rstudio.github.io/leaflet/

- https://dashboard.bpk.go.id/app/data-berbasis-wilayah

- https://dashboard.bpk.go.id/app/covid

